---
title: "CBE R Workshop"
author: "Max & Evelyn"
output:
  rmarkdown::html_document:
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## {.tabset}

### Intro to R and RMarkdown

This is what a code chunk is:

```{r}
print("hello world! :)")
```

The code within the chunk, which is demarcated by three tick symbols (`), will be executed. You can add tinker with different options in the **chunk header**, which is enclosed by the two brackets (\{\}). One useful chunk option is *echo = FALSE*, which suppresses the code from showing and only produces an output.

For example, if we add *echo = FALSE* to the above code chunk, we only get the printed output.

```{r echo=FALSE}
print("hello world! :)")
```

You can "knit" an RMarkdown by clicking the yarn icon, or by using the keyboard shortcut (there are a lot of these in RStudio): CTRL/CMD + SHIFT + K. This creates a prettier HTML output.

You can save a document by going to File-> Save or with the shortcut CTRL/CMD + S.

```{r}
print("Hello world")

# The above is a line of code. This is a comment.
```

You can create comments, or sections of text that will not be executed as code, by adding a '#' in front of the section.

### Loading packages and importing data

Different packages in R do different things. There's a package that helps you plot pretty things called **ggplot**, a package that contains a bunch of useful functions for data transformation and analysis called **dplyr**, and even a fortune teller package. There's even a package that contains the first two packages. We'll be installing this one.

```{r}
#install.packages("tidyverse")
```

Once you've installed it, you need to *load* the package before use. Every time you restart R, it forgets about your previous session. You can load packages like so:

```{r}
library(tidyverse)
```

### Exploratory data analysis

We will be working with a publicly available dataset from Our World in Data called "Coronavirus (COVID-19) Vaccinations".

Once you have it downloaded, make sure it is in the same folder as the RMarkdown document you have with code. You can read in a CSV file using a function called read.csv("FILENAME.csv")

```{r}
vaccine_data <- read_csv("vaccine_data.csv")
```

We can use the summary function to immediately display a number of summary statistics.

```{r}
summary(vaccine_data)
```

Let's say we want to take a peek at our spreadsheet, but we we're not feeling numbers today. We can use the view function to see the data right in RStudio. 

```{r eval=FALSE}
View(vaccine_data)
```

Glimpse is alternative way to view part of the data in R (note that this requires the TidyVerse library to be loaded).

```{r}
glimpse(vaccine_data)
```

You might notice symbols like <chr> and <dbl> in the above output. These are known as variable types in R. If you have learned another programming language, you might remember that you cannot add a number to a word (oftentimes this will result in *undefined behavior*), and the same is true in R. In the case of the above output <chr> stands for character, which is functionally the same as a word. You can recognize a character data type because characters are always surrounded by quotation marks. Doubles, or numerics as they are known in R, are essentially numbers. In some programming languages, integers (whole numbers), and doubles (decimal numbers) are considered as different datatypes — there is no such distinction in R. 

You might notice that some values that were printed out above said "NA" instead of a number (or a character). This often means that despite there being other values associated with a particular data entry, there was no particular value recorded for that specific part of the data entry. It would be wasteful to remove the rest of the data associated with the entry (as it could be useful for other applications), but at the same time NA values can make certain calculations difficult. For example, if we were to sum all of the daily vaccinations without any manipulation of the data, the end result is an NA value, which is completely useless. 

```{r}
sum(vaccine_data$daily_vaccinations)
```

One of the simplest ways to deal with NA values is to use the `na.omit` command, which omits any NA values from a dataset. This resolves the issue we saw above.

```{r}
sum(na.omit(vaccine_data$daily_vaccinations))
```

You can think of variables and columns as synonymous in a data table.

### Data transformation with dplyr

Now that we've checked our data for any loose ends, like NA values, let's get to manipulating the data. The **dplyr** package contains various functions that help you transform variables.

Here is a quick list of what the most common functions do:
* `filter()` - subsets your data by certain values
* `select()` - selects or deselects certain variables
* `arrange()` - sorts your data in descending or ascending value
* `summarize()` - calculates an output that is a single value
* `mutate()` - creates new columns

Let's say we're only interested in viewing the observations from February 21, 2021.

```{r}
vaccine_data %>%
  filter(date == "2021-02-21")
```

Now you can see there are only 65 observations that match this condition.

We only want the variable `people_fully_vaccinated`.

```{r}
vaccine_data %>%
  select(people_fully_vaccinated)
```

What if we want all variables *besides* `people_fully_vaccinated`?

```{r}
vaccine_data %>%
  select(-people_fully_vaccinated)
```

Now let's see which state has the highest number of `people_fully_vaccinated` as of February 21, 2021.

```{r}
vaccine_data %>%
  filter(date == "2021-02-21") %>%
  arrange(desc(people_fully_vaccinated))
```

`summarize()` is a useful function for basic statistics, like `mean()`, `median()`, or `sum()`. 

```{r}
vaccine_data %>%
  filter(date == "2021-02-21",
         location != "United States",
         !is.na(people_fully_vaccinated)) %>%
  summarize(mean(people_fully_vaccinated))
```

However, `mutate()` is useful for calculating a unique value per observation. Let's say we want to find the rate of vaccinations and distributions:

```{r}
vaccine_data %>%
  filter(!is.na(people_vaccinated),
         !is.na(total_distributed)) %>%
  mutate(newvariable = people_vaccinated / total_distributed)
```

There's so much more you can do with **dplyr**. Another useful function is *group_by()*, which creates groups based on a certain variable. For example, finding the mean number of people fully vaccinated isn't very helpful when your data contains observations from New York in December 2020 and Oregon in February 2021.

*group_by()* essentially sidesteps this problem of comparison by calculating statistics within a given group.

```{r}
vaccine_data %>%
  select(date, people_fully_vaccinated, location) %>%
  filter(!is.na(people_fully_vaccinated),
         location != "United States") %>%
  group_by(date) %>%
  summarize(mean(people_fully_vaccinated))

# Always gut check your results!
```


### Plotting with ggplot

When you have a good grasp of your findings, you might want to visualize those results. The **ggplot2* package has a robust collection of functions and themes to do this in the PRETTIEST manner possible.

We won't go over examples of each, but here's a cheat sheet of what each function does:
* `geom_bar()` creates a bar plot, which is good for graphing categorical variables.
* `geom_hist()` creates a histogram.
* `geom_point()` creates a scatter plot, good for continuous (or numerical) variables.
* `geom_boxplot()` creates a boxplot, good for looking at central tendency and distribution.
* `geom_line()` creates a line graph, good for continuous variables.

Let's say we want to see how vaccinations per capita have progressed over time in the state of Massachusetts.

```{r include=FALSE}
# We need to first prepare our data by filtering for Massachusetts only and
# selecting the variable people_vaccinated_per_hundred.

massonly <- vaccine_data %>%
  filter(location == "Massachusetts",
         !is.na(people_vaccinated_per_hundred)) %>%
  mutate(date = as.Date(date))

# By using the "<-" assignment arrow, we've stored the changes we made to the
# dataset in a new object called massonly.
```

```{r}
# First, we need to start with ggplot().
ggplot(data = massonly) +
  # Note how we're using plus signs instead of the %>% operator! This is only
  # for when using the ggplot2 package.
  geom_point(mapping = aes(x = date, y = people_vaccinated_per_hundred))
```

You can replace `geom_point()` with `geom_line()`. It's that easy to create plots using **ggplot2**.

```{r}
# Run it back
ggplot(data = massonly) +
  geom_line(mapping = aes(x = date, y = people_vaccinated_per_hundred))
```

Now that this is looking more sensible, we can add labels using the `labs()` function. You can use `xlab()` and `ylab()` to change the axis titles.

```{r}
# Make it make sense
ggplot(data = massonly) +
  geom_line(mapping = aes(x = date, y = people_vaccinated_per_hundred)) +
  labs(title = "Vaccination Progress in MA",
       subtitle = "Vaccination has picked up since mid-January",
       caption = "Source: Our World in Data") +
  xlab("Date") +
  ylab("People vaccinated per hundred")
```

You can even make it look pretty with a variety of themes.

```{r}
ggplot(data = massonly) +
  geom_line(mapping = aes(x = date, y = people_vaccinated_per_hundred)) +
  labs(title = "Vaccination Progress in MA",
       subtitle = "The progress has been fairly linear since mid-January",
       caption = "Source: Center for Disease Control") +
  xlab("Date") +
  ylab("People vaccinated per hundred") +
  # Simply add a theme.
  theme_light()
```

We won't go too in depth, but the world is your oyster when it comes to visualizing with R. Below is a neat graph that Max and I created using this dataset. Can you replicate it?

```{r echo=FALSE}
dat <- vaccine_data %>%
  filter(location %in% c("Texas", "New York State"),
         !is.na(total_distributed),
         !is.na(total_vaccinations)) %>%
  group_by(date, location) %>%
  mutate(rate = total_vaccinations / total_distributed,
         date = as.Date(date))

ggplot(data = dat, mapping = aes(x = date, y = rate, color = location)) +
  geom_line() +
  xlab("Date") +
  ylab("Rate") +
  labs(title = "Vaccination-distribution rate in NY & TX",
       caption = "Source: Our World in Data") +
  theme_minimal()
```


### Other R applications

You can do so much more with R, including but not limited to mapping / GIS stuff!!! And making a website! And election predictions, and [COVID dashboards](https://blog.rstudio.com/2020/11/19/using-shiny-in-production-to-monitor-covid-19/), and machine learning.

If you want to learn more, I highly recommend the [R primers](https://rstudio.cloud/learn/primers) created by RStudio.









