---
title: "CBE R Workshop"
author: "Max & Evelyn"
date: "2/21/2021"
output:
  rmarkdown::html_document:
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro to R and RMarkdown

This is what a code chunk is:

```{r}
print("hello world! :)")
```

The code within the chunk, which is demarcated by three tick symbols (`), will be executed. You can add tinker with different options in the **chunk header**, which is enclosed by the two brackets (\{\}). One useful chunk option is *echo = FALSE*, which suppresses the code from showing and only produces an output.

For example, if we add *echo = FALSE* to the above code chunk, we only get the printed output.

```{r echo=FALSE}
print("hello world! :)")
```

You can "knit" an RMarkdown by clicking the yarn icon, or by using the keyboard shortcut (there are a lot of these in RStudio): CTRL/CMD + SHIFT + K. This creates a prettier HTML output.

You can save a document by going to File-> Save or with the shortcut CTRL/CMD + S.

```{r}
print("Hello world")

# The above is a line of code. This is a comment.
```

You can create comments, or sections of text that will not be executed as code, by adding a '#' in front of the section.

## Loading libraries and importing data

Different libraries in R do different things. There's a library that helps you  



## Exploratory analysis & descriptive stats

First, lets load in some data. We can do this using a function called read.csv("FILENAME.csv")
```{r}
vaccine_data = read.csv("vaccine_data.csv")
```

We can use the summary function to immediately display a number of summary statistics.

```{r}
summary(vaccine_data)
```

Let's say we want to take a peek at our spreadsheet, but we we're not feeling numbers today. We can use the view function to see the data right in RStudio. 

```{r}
View(vaccine_data)
```

Summary
View
glimpse
variable types
NA values





## Data transformation with dplyr

filter
select
mutate
arrange
summarise



## Plotting with ggplot

geom_bar
geom_hist
geom_point
scatter
boxplot
facet_wrap
titles / labels
themes


scale_coord_flip
geom_hline

## Other R applications

You can do so much more with R, including but not limited to mapping / GIS stuff!!! And making a website! And election predictions, and COVID dashboards (link it), and ML

Other resources - R Primer









